Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-04 rnni_convexity.tex
pass through to Lemma 1 + rm unnecessary space (and possibly some necessary ones)

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage[misc]{ifsym}
\usepackage[foot]{amsaddr}
\usepackage[hidelinks]{hyperref}
\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
\renewbibmacro{in:}{}
\addbibresource{rnni_convexity.bib}

@@

The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
Its geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
An important property of the $\nni$ graph is that computing distance is NP-hard \autocite{Dasgupta2000-xa}.
As a result no algorithm exists to compute the distance in practical time.
A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees \autocite{...}.

More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made possible computational inference of phylogenetic time-trees \autocite{...}.
However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
\textcite{Gavryushkin2018-ol} have proposed an extension of the $\nni$ graph to the class of discrete time-trees.
The simplest such extension introduces the $\rnni$ graph on the set of ranked phylogenetic trees.
Considered with the graph-distance $\rnni$ becomes a metric space and inherits the geometric and algorithmic challenges that the $\nni$ space has been traditionally facing.
Surprisingly, most of them cannot be settled by directly translating results or applying techniques developed for $\nni$ \autocite{Gavryushkin2018-ol}.

@@

In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
Specifically, in this paper we establish the exact radius and diameter of the space.
We also show that the subset of caterpillar trees is convex in $\rnni$.
Moreover, we can provide an approximate algorithm that computes exact distances for small trees.
The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.

In the rest of this chapter we formally introduce non-standard definitions necessary for this paper.

A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all leaves of $T$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
The latter means that if $u$ and $v$ are two internal nodes of $T$ such that there exists a path from a taxon $x \in X$ to the root which first passes through $u$ and then through $v$ then $\rank(u) < \rank(v)$.
Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
Since all trees in this paper are ranked trees, we will abuse the notation and drop the ranking from the notation.
We will also refer to a ranked tree simply by tree.
For a tree $T$, we will use $\rank_T$ to refer to its ranking.
A ranked tree $T$ without its leaf labels will be called \emph{tree topology}.

Our definition of a ranked tree implies a natural notion of the edge length.:
A tree edge $(u,v)$ of $T$ has length $l$ if $|\rank_T(v) - \rank_T(u)| = l$.

@@

\todo{LC: Explain structure of paper}

@@

Within this section we will at first described how we can compute exact distances between trees in the $\rnni$ graph (Section~\ref{section:alg_RNNI_graph}) by using results of \autocite{Gavryushkin2018-ol}.
Afterwards, we introduce two algorithms, namely $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
The first one takes arbitrary trees as input while the latter one is restricted to trees with a certain tree topology.
We will use them later to prove some geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
Additionally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree.
What we mean by far away depends on the input tree and will be discussed after introducing this algorithm in Section~\ref{section:alg_mdtree}.

@@

In Section 3.3 of \autocite{Gavryushkin2018-ol} an exact algorithm for computing the $\rnni$ graph has been introduced.
We implemented this algorithm and used it for computing this graph for trees on up to seven taxa.
\todo{LC: $7$ or $8$? - I'll need to check this}
Once the graph is computed, one can get distances by using algorithms such as Floyd-Warshall \autocite{Floyd1962-ew} or Dijkstra \autocite{Dijkstra1959-ph}.
However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we already run out of space when we want to save all pairwise distances for trees on seven taxa.
Therefore, we compute distances between certain pairs of trees and use the symmetry of $\rnni$ to draw conclusions about the remaining distances.

@@

Before introducing the algorithm we need the following definitions.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
Note that the list of all non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
For the rest of this paper we will use the notion cluster and mean non-trivial clusters only because all trees on $n$ taxa share the trivial clusters.
We will refer to this list of clusters $n-2$ of a given tree on $n$ taxa as \emph{cluster representation} of the tree.

We are now presenting $\findpath$ (Algorithm~\ref{alg:find_path}), an algorithm for computing paths between any two given trees in $\rnni$.
Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the destination tree).
Then $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
$\findpath$ computes a path $p$ which initially consists of $T$ only and is extended stepwise.
In order to do this, the clusters $C_1, \ldots, C_{n-2}$ of $R$ are considered in that order.
Let $T'$ be the last tree of $p$ when arriving at cluster $C_k$.
We assume that the clusters $C_1, \ldots, C_{k-1}$ are induced by nodes of rank $1, \ldots, k-1$ in $T'$, respectively.
In the first step it is $T' = T$ and $C_k = C_1$.
The aim is now to find a sequence of $\rnni$ moves, starting at $T'$, that do not change the clusters $C_1, \ldots, C_{k-1}$, but result in a tree containing clusters $C_1, \ldots, C_{k}$.
Therefore, we need to find the smallest cluster containing all elements of $C_k$ in $T'$.
This node is denoted by $\mrca_{T'}(C_k)$ (short for \emph{most recent common ancestor}).
The rank of $\mrca_{T'}(C_k)$ in $T'$ is higher than the rank of the node inducing $C_k$ in $R$, because $C_1, \ldots, C_{k-1}$ are already present in $T'$ and we assume that $C_k$ is not.
Now perform an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
With Lemma~\ref{lemma:mrca_move} we know that such an $\rnni$ move always exists if long as $T'$ does not induce the cluster $C_k$.
So we can perform such moves until a tree $T''$ with $\mrca_{T''}(C_k)$ is reached that contains a node inducing $C_k$ that has the same rank $k$ in $T''$ as in $R$.
Appending all trees on this path from $T'$ to $T''$ and repeating this procedure for all clusters of $R$ results in a path $p$ from $T$ to $R$.

@@

\begin{proof}

    Let $T, R$ and $T'$ be trees as defined in the lemma and $[C_1, \ldots, C_{n-2}]$ the cluster representation of $R$.
    We want to prove that there is exactly one $\rnni$ move on $T'$ that results in a tree $T''$ where $\mrca_{T''}(C_k) < \mrca_{T'}(C_k)$.
    Notice, that the only $\rnni$ move on $T$ that could decrease the rank of $\mrca_{T'}(C_k)$ are $\nni$ movs on the edge between $\mrca_{T'}(C_k)$ and its child (if this edge has length one), or a rank move between $\mrca_{T'}(C_k)$ and the node with rank one less.
    Therefore, we will now distinguish these two cases:
    Either there is an edge $(v,w)$ between nodes $v = \mrca_{T'}(C_k)$ and $w$ with rank $rank_{T'}(w) = rank_{T'}(v) - 1$, or there is no such edge.

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage[misc]{ifsym}
\usepackage[foot]{amsaddr}
\usepackage[hidelinks]{hyperref}
\usepackage{csquotes}
\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
\renewbibmacro{in:}{}
\addbibresource{rnni_convexity.bib}

@@

The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
Its geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
An important property of the $\nni$ graph is that computing distance is NP-hard \autocite{Dasgupta2000-xa}.
As a result no algorithm exists to compute the distance in practical time.
A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees
\todo{cite Chris Whidden}
\autocite{}.

More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made possible computational inference of phylogenetic time-trees
\todo{cite MrBayes, BEAST, NextStrain}
\autocite{}.
However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
These differences motivated \textcite{Gavryushkin2018-ol} to propose an extension of the $\nni$ graph to the class of discrete time-trees.
The simplest such extension introduces the $\rnni$ graph on the set of ranked phylogenetic trees.
Considered with the graph-distance $\rnni$ becomes a metric space and inherits the geometric and algorithmic challenges that the $\nni$ space has been traditionally facing.
Surprisingly, most of them cannot be settled by directly translating results or applying techniques developed for $\nni$ \autocite{Gavryushkin2018-ol}.

@@

In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
Specifically, in this paper we establish the exact radius and diameter of the space.
We also show that the subset of caterpillar trees is convex in $\rnni$, thus settling one of the open problems proposed in \autocite{Gavryushkin2018-ol}.
Furthermore, we provide an approximate algorithm that computes exact distances for small trees.
The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.

In the rest of this chapter we formally introduce the terminology used in this paper.

A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function that maps all leaves of $T$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order on the nodes given by the tree.
The latter means that if $u$ and $v$ are two internal nodes of $T$ such that there exists a path from a taxon $x \in X$ to the root which first passes through $u$ and then through $v$ then $\rank(u) < \rank(v)$.
Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
Since all trees in this paper are ranked trees, we will abuse the notation and drop the rank function from the notation.
We will also be simply saying tress to mean ranked phylogenetic trees.
For a tree $T$, we will use $\rank_T$ to refer to its ranking.
A ranked tree $T$ without its leaf labels will be called \emph{tree topology}.

Our definition of a ranked tree implies a natural notion of the edge length -- we call $|\rank_T(v) - \rank_T(u)|$ the length of an edge $(u, v)$.

@@

\todo[inline]{LC: Explain structure of paper}

@@

We start this section with a discussion of the algorithm proposed in \autocite{Gavryushkin2018-ol} to compute the exact $\rnni$ distance between trees by pre-computing the $\rnni$ graph.
Then we introduce two algorithms, $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
The first of these algorithms takes arbitrary trees as an input while the latter is restricted to trees with a certain tree topology.
Both of these algorithms will become an important ingredient to establish several geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
Finally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree (see Section~\ref{section:alg_mdtree} for details).
While this algorithm is interesting on its own (e.g.\ for simulation studies), we will again apply the algorithm to study the geometry of the $\rnni$ space.

@@

\textcite[Section 3.3]{Gavryushkin2018-ol} constructed an exact algorithm for computing the $\rnni$ graph.
\todo{Do we make use of tanglegrams?}
\todo{GitHub link?}
We implemented this algorithm and used it for computing this graph for trees on up to seven taxa.
\todo{LC: $7$ or $8$? - I'll need to check this}
Once the graph is computed, one can get distances by using algorithms such as Floyd-Warshall \autocite{Floyd1962-ew} or Dijkstra \autocite{Dijkstra1959-ph}.
However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we already run out of space when we want to save all pairwise distances for trees on seven taxa.
\todo{Is this the hint we're using tanglegrams?
Either way this statement is too vague and requires more detail.
Also, ``run out of space'' should be a bit less frank.
Finally, think whether you want single out this paragraph as a special subsection?
Either way you should say why you need this implementation.}
Therefore, we compute distances between certain pairs of trees and use the symmetry of $\rnni$ to draw conclusions about the remaining distances.

@@

Before introducing the algorithm we need the following definitions.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
As all trees on $n$ taxa share all trivial clusters, we will use the notion cluster to mean non-trivial clusters.
Note that the list of all non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
We will refer to this list of $n-2$ clusters for a given tree on $n$ taxa as a \emph{cluster representation} of the tree.

We are now ready to introduce $\findpath$ (Algorithm~\ref{alg:find_path}), an efficient algorithm for computing paths between any two given trees $T$ and $R$ in $\rnni$.
Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
Then $C_i$ is induced by the node of rank $i$ for $i = 1, \ldots, n-2$.
$\findpath$ computes a path $p$ which initially consists of $T$ only and is extended stepwise.
In order to do this, the clusters $C_1, \ldots, C_{n-2}$ of $R$ are considered in their order.
Let $T'$ be the last tree of $p$ when arriving at cluster $C_k$.
We assume that the clusters $C_1, \ldots, C_{k-1}$ are induced by nodes of rank $1, \ldots, k-1$ in $T'$, respectively.
In the first step it is $T' = T$ and $C_k = C_1$.
The aim is now to find a sequence of $\rnni$ moves, starting at $T'$, that do not change the clusters $C_1, \ldots, C_{k-1}$, but result in a tree containing clusters $C_1, \ldots, C_{k}$.
Therefore, we need to find the smallest cluster containing all elements of $C_k$ in $T'$.
This node is denoted by $\mrca_{T'}(C_k)$ (short for \emph{most recent common ancestor}).
The rank of $\mrca_{T'}(C_k)$ in $T'$ is higher than the rank of the node inducing $C_k$ in $R$, because $C_1, \ldots, C_{k-1}$ are already present in $T'$ and we assume that $C_k$ is not.
Now perform an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
\todo{We haven't seen Lemma 1 yet.}
With Lemma~\ref{lemma:mrca_move} we know that such an $\rnni$ move always exists as long as $T'$ does not induce the cluster $C_k$.
So we can perform such moves until a tree $T''$ is reached with $\mrca_{T''}(C_k)$ inducing $C_k$ and having has the same rank $k$ in $T''$ as in $R$.
Appending all trees on this path from $T'$ to $T''$ and repeating this procedure for all clusters of $R$ results in a path $p$ from $T$ to $R$.

@@

\begin{proof}
Let $T, R$ and $T'$ be trees as defined in the lemma and $[C_1, \ldots, C_{n-2}]$ the cluster representation of $R$.
We want to prove that there is exactly one $\rnni$ move on $T'$ that results in a tree $T''$ where $\mrca_{T''}(C_k) < \mrca_{T'}(C_k)$.
Notice, that the only $\rnni$ move on $T$ that could decrease the rank of $\mrca_{T'}(C_k)$ are $\nni$ movs on the edge between $\mrca_{T'}(C_k)$ and its child (if this edge has length one), or a rank move between $\mrca_{T'}(C_k)$ and the node with rank one less.
Therefore, we will now distinguish these two cases:
Either there is an edge $(v,w)$ between nodes $v = \mrca_{T'}(C_k)$ and $w$ with rank $rank_{T'}(w) = rank_{T'}(v) - 1$, or there is no such edge.
